A free/open-source Kazakh-Tatar machine translation system

### 2021-05-27 paper/paper.tex
OFTC, http -> https, url fixups

This paper presents a prototype shallow-transfer rule-based machine translation
system between Kazakh and Tatar.

The paper will be laid out as follows: Section~\ref{sec:prev}\ gives a short review of some previous work in the area of Turkic--Turkic language machine translation; Section~\ref{sec:lang} introduces Kazakh and Tatar and compares their grammar; 
Section~\ref{sec:sys}\ describes the system and the tools used to construct it;
Section~\ref{sec:eval}\ gives a preliminary evaluation of the system; and
finally Section~\ref{sec:conc}\ describes our aims for future work and some
concluding remarks.

@@

An MT system between Tatar and Kazakh is potentially of great use to the language communities.  Direct MT can save time and money over going through e.g.\ Russian (and the system is much easier to develop).  Some accompanying language resources (e.g.,\ morphol.\ analysers, disambiguators) can also be repurposed.

@@

The past tense systems of Kazakh and Tatar have a many-to-many correspondence.  As shown 
in Table~\ref{tab:pasttense}, at a basic level, in the past tense, Kazakh 
differentiates [±eyewitness]\footnote{``Eyewitness'' is a convenient term for this feature, though it 
may be better expressed as simply ``reliability of knowledge'' (which indeed often equates to whether the 
knowledge was acquired first-hand or not) in many cases.} (where [-eyewitness] is used for cases of both 
potentially unreliable information and newly discovered information) and [±recent], whereas Tatar has only 
three categories: eyewitness, non-eyewitness, and newly-acquired-information---all with no [±recent] distinction.  As 
an example of the many-to-many correspondence that this results in, Tatar has a single non-eyewitness 
past tense morpheme (\texttt{-GAн-}) while Kazakh has a recent non-eyewitness past (\texttt{-Iп-}) and a 
distant non-eyewitness past (\texttt{-GAн екен-}).  On the other hand, these two non-eyewitness past 
forms in Kazakh are used for both potentially unreliable information and newly acquired information, whereas 
in Tatar, non-eyewitness (\texttt{-GAн-}) and newly-acquired-information (\texttt{-GAн- икән}) past forms are distinguished.

\begin{table}[htbp]
	\centering
	\subfloat[Kazakh past tense morphology]{
		\begin{tabular}{lll}
			\toprule
			 & [$+$recent] & [$-$recent] \\\midrule
			{}[$+$reliable] & -DI- & -GAн- \\
			{}[$-$reliable] & -Iп- & -GAн екен- \\ \bottomrule
		\end{tabular}
%		\begin{tabular}{lll}
%			\toprule
%			 & recent & non-recent \\\midrule
%			eyewitness/reliable & -DI- & -GAн- \\
%			non-eyewitness & \multirow{2}*{-Iп-} & \multirow{2}*{-GAн екен-} \\
%			{} / newly acquired information & & \\ \bottomrule
%		\end{tabular}

	}\\
	\subfloat[Tatar past tense morphology]{
		\begin{tabular}{lll}
			\toprule
			 & [$-$new] & [$+$new] \\\midrule
			{}[$+$reliable] & -DI- & -GAн- икән \\
			{}[$-$reliable] & -GAн- & --- \\\bottomrule
		\end{tabular}
%		\begin{tabular}{lll}
%			\toprule
%			 & past \\\midrule
%			eyewitness & -DI- \\
%			non-eyewitness/hearsay & -GAн- \\
%			newly acquired information & -GAн- икән \\\bottomrule
%		\end{tabular}
	}
	\caption{A comparison of the basic past-tense morphology of Kazakh and Tatar}
	\label{tab:pasttense}
\end{table}
\vspace{-1ex}
%\vspace{-1ex}
The morphotactics of the cognate Kazakh 
distant non-eywitness past (\texttt{-GAн екен-}) and Tatar newly-acquired-information past (\texttt{-GAн- икән}) are 
different.\footnote{This comparison is made without regard to semantic alignment.}  Specifically, in both languages, the person agreement takes the form of a person copula suffix; however
in Kazakh this suffix follows the tense morphemes (e.g., барған екенсің ``you apparently went''), whereas in 
Tatar this suffix intervenes between the two pieces of the `compound' tense morpheme (e.g., баргансың икән ``I guess you went'').
% I think that the meaning of the Tatar "баргансың икән" is also "It turns out you went, It appears that you went,
% They say you went", i.e. it is closer to Kazakh. "I guess you went" sounds like you are supposing that, which is not the case.
% Maybe I just don't know the whole semantics of the english "to guess". If you are sure about it, just leave it as it is.

@@

%2 or 3 specific examples
% * Different case governance of postpositions (grep for WARNINGs in bidix)
% * музейлік экспозиция > музей экспозициясе; экспозициялық зал > экспозиция залы
% ** This works, there is a rule for that: if an /adjective + noun/ is translated with /n.attr + noun/, <px3sp> affix is added in Tatar 
% * Мен оқуды жақсы көремін > Мин укырга яратам.
% Also, see this page: http://wiki.apertium.org/wiki/Kazakh_and_Tatar/Pending_tests
There are a number of minor syntactic differences in Tatar and Kazakh, which include differences in verb valencies in equivalent translations, as well Tatar's reliance on a ``true'' infinitive that is used in place of various verbal noun and verb adverb forms in Kazakh.

@@

The system is based on the Apertium machine translation 
platform \citep{apertium/2011}.\footnote{\url{http://www.apertium.org}} The 
platform was originally aimed at the Romance languages of the Iberian peninsula, but has also been adapted for 
other, more distantly related, language pairs.
The whole platform, both programs and data, are licensed under the Free Software Foundation's General Public 
Licence\footnote{\url{http://www.fsf.org/licensing/licenses/gpl.html}} (GPL) and all the software and data for the 
30 supported language pairs (and the other pairs being worked on) is available for download from the project 
website.

@@

The output of each morphological analyser is highly ambiguous, measured at around 2.4 morphological analyses per form for Kazakh.  The goal of the CG rules is to select the correct analysis when there are multiple analyses.  Currently, ambiguity is down to 1.4 analyses per form.
%As of r43664, the ambiguity was down to 1.4 analyses per form.

One reason for the still high level of morphological ambiguity is a series of affixes in both languages which can each be analysed as some combination of verbal nouns, verbal adjectives, substantitivised verbal adjectives (i.e., verbal adjectives with a null modified noun), and even finite forms.\footnote{Despite the fact that the various suffixes in this category pattern differently and do not form a single natural class, most grammars of the languages label all them as simply ``gerunds'' or ``participles''.}

@@

\begin{table}[htbp]
	\centering
	\subfloat[Naïve coverage of the Kazakh-Tatar direction]{
		\begin{tabular}{llrrr}
			\toprule
			Corpus          & Tokens & Coverage & stdev\\
			\midrule
			RFERL 2010      & 3.2M   & 90.19\% & $\pm$ 0.23\% \\
			RFERL 2012      & 2.9M   & 89.74\% & $\pm$ 0.59\% \\
			Wikipedia       & 1.2M   & 80.75\% & $\pm$ 5.23\% \\
			\bottomrule
		\end{tabular}
	}\\
	\subfloat[Naïve coverage of the Tatar-Kazakh direction]{
		\begin{tabular}{llrrr}
			\toprule
			Corpus          & Tokens & Coverage & stdev\\
			\midrule
			RFERL 2007-'12   & 1.2M   & 82.24\% & $\pm$ 2.88\% \\
			New Testament   & 137K   & 91.79\% & $\pm$ 1.39\% \\
			Wikipedia       & 128K   & 81.36\% & $\pm$ 1.48\% \\
			\bottomrule
		\end{tabular}
	}

@@

%FIXME NOW
All evaluation was tested against version 0.1.0, or {\tt {\small r45441}} in the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}
%The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}

Lexical coverage of the system is calculated over freely available corpora of Kazakh and Tatar.

For Kazakh, two years worth of content (2010 and 2012) from Radio Free Europe / Radio Liberty (RFERL)'s Kazakh-language service,\footnote{\url{http://www.azattyq.org/}} as well as a recent dump of Wikipedia's articles in Kazakh\footnote{\url{http://kk.wikipedia.org/}; {\smallertt kkwiki-20130408-pages-articles.xml.bz2}} were used.

For Tatar, a dump of articles from the Tatar Wikipedia,\footnote{\url{http://tt.wikipedia.org/}; {\smallertt ttwiki-20130205-pages-articles.xml.bz2}} a translation of the New Testament, and content from RFERL's Tatar-language service\footnote{\url{http://www.azatlyk.org/}} from early 2007 to early 2012 were used for testing.

@@

We plan to continue development on the pair; the coverage of the system is already quite high, and although we intend to
increase it to 95\% on the corpora we have, the main work will be improving the quality of translation by adding
more rules, starting with the CG module.
The long-term plan is to integrate the data created with other open-source data for Turkic languages in order to make
transfer systems between all the Turkic language pairs.  Related work is currently ongoing with Chuvash--Turkish and Turkish--Kyrgyz.

The system is available as free/open-source software under the GNU GPL and the whole system may be downloaded
from SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}

>>>>

This paper presents a prototype shallow-transfer rule-based machine translation
system between Kazakh and Tatar.
The paper will be laid out as follows: Section~\ref{sec:prev}\ gives a short review of some previous work in the area of Turkic--Turkic language machine translation; Section~\ref{sec:lang} introduces Kazakh and Tatar and compares their grammar; 
Section~\ref{sec:sys}\ describes the system and the tools used to construct it;
Section~\ref{sec:eval}\ gives a preliminary evaluation of the system; and
finally Section~\ref{sec:conc}\ describes our aims for future work and some
concluding remarks.

@@

An MT system between Tatar and Kazakh is potentially of great use to the language communities.  Direct MT can save time and money over going through e.g.\ Russian (and the system is much easier to develop).  Some accompanying language resources (e.g.,\ morphololgical analysers, disambiguators) can also be repurposed.

@@

The past tense systems of Kazakh and Tatar have a many-to-many correspondence.  As shown 
in Table~\ref{tab:pasttense}, at a basic level, in the past tense, Kazakh 
differentiates [±eyewitness]\footnote{``Eyewitness'' is a convenient term for this feature, though it 
may be better expressed as simply ``reliability of knowledge'' (which indeed often equates to whether the 
knowledge was acquired first-hand or not) in many cases.} (where [-eyewitness] is used for cases of both 
potentially unreliable information and newly discovered information) and [±recent], whereas Tatar has only 
three categories: eyewitness, non-eyewitness, and newly-acquired information---all with no [±recent] distinction.  As 
an example of the many-to-many correspondence that this results in, Tatar has a single non-eyewitness 
past tense morpheme (\texttt{-GAн-}) while Kazakh has a recent non-eyewitness past (\texttt{-Iп-}) and a 
distant non-eyewitness past (\texttt{-GAн екен-}).  On the other hand, these two non-eyewitness past 
forms in Kazakh are used for both potentially unreliable information and newly acquired information, whereas 
in Tatar, non-eyewitness (\texttt{-GAн-}) and newly-acquired information (\texttt{-GAн- икән}) past forms are distinguished.

\begin{table}[htbp]
	\centering
	\subfloat[Kazakh past tense morphology]{
		\begin{tabular}{lll}
			\toprule
			 & \texttt{[+recent]} & \texttt{[-recent]} \\\midrule
			{}\texttt{[+reliable]} & -DI- & -GAн- \\
			{}\texttt{[-reliable]} & -Iп- & -GAн екен- \\ \bottomrule
		\end{tabular}
%		\begin{tabular}{lll}
%			\toprule
%			 & recent & non-recent \\\midrule
%			eyewitness/reliable & -DI- & -GAн- \\
%			non-eyewitness & \multirow{2}*{-Iп-} & \multirow{2}*{-GAн екен-} \\
%			{} / newly acquired information & & \\ \bottomrule
%		\end{tabular}

	}\\
	\subfloat[Tatar past tense morphology]{
		\begin{tabular}{lll}
			\toprule
			 & \texttt{[-new]} & \texttt{[+new]} \\\midrule
			{}\texttt{[+reliable]} & -DI- & -GAн- икән \\
			{}\texttt{[-reliable]} & -GAн- & --- \\\bottomrule
		\end{tabular}
%		\begin{tabular}{lll}
%			\toprule
%			 & past \\\midrule
%			eyewitness & -DI- \\
%			non-eyewitness/hearsay & -GAн- \\
%			newly acquired information & -GAн- икән \\\bottomrule
%		\end{tabular}
	}
	\caption{A comparison of the basic past-tense morphology of Kazakh and Tatar}
	\label{tab:pasttense}
\end{table}
\vspace{-1ex}
%\vspace{-1ex}
The morphotactics of the cognate Kazakh 
distant non-eywitness past (\texttt{-GAн екен-}) and Tatar newly-acquired-information past (\texttt{-GAн- икән}) are 
different.\footnote{This comparison is made without regard to semantic alignment.}  Specifically, in both languages, the person agreement takes the form of a person copula suffix; however
in Kazakh this suffix follows the tense morphemes (e.g., барған екенсің ``you apparently went''), whereas in 
Tatar this suffix intervenes between the two pieces of the `compound' tense morpheme (e.g., баргансың икән ``I guess you went'').
% I think that the meaning of the Tatar "баргансың икән" is also "It turns out you went, It appears that you went,
% They say you went", i.e. it is closer to Kazakh. "I guess you went" sounds like you are supposing that, which is not the case.
% Maybe I just don't know the whole semantics of the english "to guess". If you are sure about it, just leave it as it is.

@@

%2 or 3 specific examples
% * Different case governance of postpositions (grep for WARNINGs in bidix)
% * музейлік экспозиция > музей экспозициясе; экспозициялық зал > экспозиция залы
% ** This works, there is a rule for that: if an /adjective + noun/ is translated with /n.attr + noun/, <px3sp> affix is added in Tatar 
% * Мен оқуды жақсы көремін > Мин укырга яратам.
% Also, see this page: https://wiki.apertium.org/wiki/Kazakh_and_Tatar/Pending_tests
There are a number of minor syntactic differences in Tatar and Kazakh, which include differences in verb valencies in equivalent translations, as well Tatar's reliance on a ``true'' infinitive that is used in place of various verbal noun and verb adverb forms in Kazakh.

@@

The system is based on the Apertium machine translation 
platform \citep{apertium/2011}.\footnote{\url{http://apertium.org}} The 
platform was originally aimed at the Romance languages of the Iberian peninsula, but has also been adapted for 
other, more distantly related, language pairs.
The whole platform, both programs and data, are licensed under the Free Software Foundation's General Public 
Licence\footnote{\url{http://www.fsf.org/licensing/licenses/gpl.html}} (GPL) and all the software and data for the 
30 supported language pairs (and the other pairs being worked on) is available for download from the project 
website.

@@

The output of each morphological analyser is highly ambiguous, measured at around 2.4 morphological analyses per form for Kazakh and 2.1 for Tatar.  The goal of the CG rules is to select the correct analysis when there are multiple analyses.  Currently, ambiguity is down to 1.4 analyses per form for Kazakh and 1.7 for Tatar.\footnote{The Tatar CG is largely based on rules generalised from Kazakh's CG, but it has not received as much attention.}
%As of r43664, the ambiguity was down to 1.4 analyses per form.

One reason for the still high level of ambiguity is a series of affixes in both languages which can each be analysed variously as verbal nouns, verbal adjectives, substantitivised verbal adjectives (verbal adjectives with a null modified noun), and even finite forms.\footnote{Despite the fact that the various suffixes in this category pattern differently and do not form a single natural class, most grammars of the languages label all them as simply ``gerunds'' or ``participles''.}

@@

\begin{table}[htbp]
	\centering
	\subfloat[Naïve coverage of the Kazakh-Tatar direction]{
		\begin{tabular}{lrrr}
			\toprule
			Corpus          & Tokens & Coverage & stdev\\
			\midrule
			RFERL 2010      & 3.2M   & 90.19\% & $\pm$ 0.23\% \\
			RFERL 2012      & 2.9M   & 89.74\% & $\pm$ 0.59\% \\
			Wikipedia       & 1.2M   & 80.75\% & $\pm$ 5.23\% \\
			\bottomrule
		\end{tabular}
	}\\
	\subfloat[Naïve coverage of the Tatar-Kazakh direction]{
		\begin{tabular}{lrrr}
			\toprule
			Corpus          & Tokens & Coverage & stdev\\
			\midrule
			RFERL 2007-'12   & 1.2M   & 82.24\% & $\pm$ 2.88\% \\
			New Testament   & 137K   & 91.79\% & $\pm$ 1.39\% \\
			Wikipedia       & 128K   & 81.36\% & $\pm$ 1.48\% \\
			\bottomrule
		\end{tabular}
	}

@@

%FIXME NOW
All evaluation was tested against version 0.2.0, or {\tt {\small r45482}} in the Apertium SVN.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-kaz-tat}}
%The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}

Lexical coverage of the system is calculated over freely available corpora of Kazakh and Tatar. For Kazakh, two years worth of content (2010 and 2012) from Radio Free Europe / Radio Liberty (RFERL)'s Kazakh-language service,\footnote{\url{http://www.azattyq.org/}} as well as a recent dump of Wikipedia's articles in Kazakh\footnote{\url{http://kk.wikipedia.org/}; {\smallertt kkwiki-20130408-pages-articles.xml.bz2}} were used.
For Tatar, a dump of articles from the Tatar Wikipedia,\footnote{\url{http://tt.wikipedia.org/}; {\smallertt ttwiki-20130205-pages-articles.xml.bz2}} a translation of the New Testament, and content from RFERL's Tatar-language service\footnote{\url{http://www.azatlyk.org/}} from early 2007 to early 2012 were used for testing.

@@

We plan to continue development on the pair; the coverage of the system is already quite high, although we intend to
increase it to 95\% on the corpora we have --- we estimate that this will mean adding around 5,000 new stems and take 1--2 months. The remaining work will be improving the quality of translation by adding
more rules, starting with the CG module.
The long-term plan is to integrate the data created with other open-source data for Turkic languages in order to make
transfer systems between all the Turkic language pairs.  Related work is currently ongoing with Chuvash--Turkish and Turkish--Kyrgyz.

The system is available as free/open-source software under the GNU GPL and the whole system may be downloaded
from sourceforge.\footnote{\url{http://sourceforge.net/projects/apertium/files/apertium-kaz-tat/}}
